##NOTES-BEGINNING
Always open up from files --> Rproj file to access git
install.packages(c("cowplot", "gridExtra"))
L"^-1*" <-- superscript notation
##NOTES-END
---
title: "4: Physical Properties of Lakes"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Investigate the relationships between temperature, light, and oxygen in lakes by analyzing monitoring data
2. Analyze changes in lake surface temperatures over time
3. Apply data analytics skills to questions about physical properties of lakes
4. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
We analyzed temperature profiles and seasonal changes on Wednesday. What were some key takeaway points about thermal stratification and mixing?
- dimictic mixing (spring & fall) -- small density gradient & wind 
- 4 degC is point at which liquid water is most dense (before prepares to become ice at lower temps)
- liquid water is geometric / crystalline so less dense 
## Session Set Up
```{r, message = FALSE}
getwd()

library(tidyverse)
library(gridExtra)
library(cowplot)
library(lubridate)

#lubridate:
#- day, month, year
#- time intervals -- to select only the hour 

NTLdata <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

theme_set(theme_classic())

```

## Data Wrangling and Exploration

### Investigate structure of dataset

```{r}
# Change sampledate to Date
NTLdata$sampledate <- as.Date(NTLdata$sampledate, "%m/%d/%y") #sample date is the name of the column... as.Date tells R to interpret as Date

# Make three data frames, one for each lake
Pauldata <- filter(NTLdata, lakename == "Paul Lake")
Peterdata <- filter(NTLdata, lakename == "Peter Lake")
Tuesdaydata <- filter(NTLdata, lakename == "Tuesday Lake")

# Make a data frame to inspect 2016 data for Paul Lake
Pauldata.2016 <- filter(Pauldata, year4 == 2016)

```

### Data Visualization

Let's look at profiles of temperature, irradiance, and dissolved oxygen over the course of a season. We will use the `plot_grid` function from the cowplot package to put the graphs together on one plot. Pay attention to the differences among the formats of each plot and how these work together in the final plot. 

```{r}
Tempprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)") +
  theme(legend.position = "none")
print(Tempprofiles2016)

Lightprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = irradianceWater, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") + #c means continuous color palette ... 0.8 is specifying start/end point so yellow not used
  labs(x = expression("Irradiance (W m"^-2*")"), y = "Depth (m)") +
  theme(legend.position = "none", 
        axis.text.y = element_blank(), axis.title.y = element_blank())
print(Lightprofiles2016)

Oxygenprofiles2016 <- 
  ggplot(Pauldata.2016, aes(x = dissolvedOxygen, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() +
  scale_x_continuous(position = "top") +
  scale_color_viridis_c(end = 0.8, option = "magma") +
  labs(x = expression("DO (mg L"^-1*")"), y = "Depth (m)", 
       color = "Julian Day") +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank())
print(Oxygenprofiles2016)

Physicalprofiles2016 <- 
  plot_grid(Tempprofiles2016, Lightprofiles2016, Oxygenprofiles2016, 
            ncol = 3, rel_widths = c(1.25, 1, 1.5))
print(Physicalprofiles2016)
#plot_grid puts together different plots 
#facet has multiple plots on same graph but done in a different way 
#can streatmline using one legend 
#makes sense to share one y-axis because plots have same y 
```

How do temperature profiles vary over the course of a season? What are the mechanisms for this pattern?

> temperature increases throughout the season 
greater sunlight radiation causes an increase in surface water temperature

How do irradiance profiles vary over the course of a season? What are the mechanisms for this pattern?

> irradiance first increases and then decreases throughout the season
irradiance is greatest in mid-June
irradiance is greater at surface
light is scattering off water molecules 
algal blooms absorb & scatter light -- more biomass at top of water column will cause light to decrease more rapidly as you go farther down
dissolved organic matter (tea colored water) can scatter light 

How do dissolved oxygen profiles vary over the course of a season? What are the mechanisms for this pattern?

> dissolved oxygen decreases throughout the season
as air temperatures and subsequently surface water temperatures increase, the water molecules move more rapidly, and some oxygen is transported into the air as water evaporates 
the water has a smaller capacity to retain the oxygen
colder water has the capacity to hold more dissolved gasses (at saturation)

>what can impact oxygen? 
1) exchange with atmosphere
2) photosynthesis 
3) respiration 
deep chlorophyll maximum in lakes ... when there is just enough light AND nutrient rich water at top of hypolimnion bottom of metalimnion 
sitting on density gradient no buoyancy issues
organisms respiring in hypolimnion -- bacterial respiration causes sudden decrease in DO in area where there is NO oxygen coming in (no light penetration for phytoplankton to grow)


## Data Analysis

### Long-term change in temperature
How much have surface temperatures increased over the period of study? Is this increase significant? Isolate surface depths and run the test for May, June, July, and August temperatures separately. Use a linear regression with year as the predictor variable and temperature as the response variable.

Steps: 

1. Add a column named "Month" to the data frame (hint: lubridate package)
2. Filter your data frame so that it only contains surface depths and months 5-8
3. Create 4 separate data frames, one for each month
4. Run a linear regression for each data frame (see instructions above)
5. For significant trends, calculate how many degrees the lake has warmed over the period of study
5. Create a ggplot of surface temps by date and facet by month. Add `geom_smooth(se = FALSE, method = lm)` to the plot to display the regression line.

```{r}
Pauldata <- Pauldata %>%
  mutate(Month = month(sampledate)) %>%
  filter(depth == 0 & Month < 11 & Month > 4)

Pauldata.May <- filter(Pauldata.test, Month == 5)
Pauldata.June <- filter(Pauldata.test, Month == 6)
Pauldata.July <- filter(Pauldata.test, Month == 7)
Pauldata.August <- filter(Pauldata.test, Month == 8)

Maytest <- lm(data = Pauldata.May, temperature_C ~ year4)
summary(Maytest)

Junetest <- lm(data = Pauldata.June, temperature_C ~ year4)
summary(Junetest)

Julytest <- lm(data = Pauldata.July, temperature_C ~ year4)
summary(Julytest)
#Temperature increases 0.06 degrees per year 
0.06*33
#1.98 degree increase over period of study for July 

Augusttest <- lm(data = Pauldata.August, temperature_C ~ year4)
summary(Augusttest)
#Temperature increases 0.04 degrees per year
0.04*33
# 1.32 degree increase over period of study for August

Tempchange.plot <-
  ggplot(Pauldata, aes(x = sampledate, y = temperature_C)) +
  geom_point() +
  geom_smooth(se = FALSE, method = lm) +
  facet_grid(rows = vars(Month))
print(Tempchange.plot)
```

Summarize your results below. How would you explain your findings and the importance of these findings to a non-specialist audience while still employing the data and statistical results?

>

## Closing Discussion
How do warming surface waters impact the physical properties of a lake? How might this affect the ecology of the lake?

How do we reconcile large amounts of variability in long-term records?
