#NOTES
There is a "Help" tab next to "Files", "Plots", "Packages", etc.
It has a search bar
#NOTES

---
title: "5: Physical Properties of Rivers"
author: "Hydrologic Data Analysis | Kateri Salk"
date: "Fall 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Analyze seasonal and interannual characteristics of stream discharge 
2. Compare discharge patterns in different regions of the United States
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion
What are the physical properties of streams and rivers? What variables might be of interest?
- discharge (volume / time) <-- relies on geometry & cross-sectional area
- flow velocity (distance / time) <-- simply horizontal 
- bed material --> erosion --> turbidity  
- mixing --> turbulent & laminar flow 
- temperature 
- slope

## Session Set Up

```{r, message = FALSE}
getwd()

library(tidyverse)
library(dataRetrieval)
library(cowplot)
library(lubridate)

theme_set(theme_classic())
```

## Discharge

One of the most important physical characteristics of a stream or river is **discharge**, the volume of water moving through the stream over a given amount of time. Discharge can be measured directly by measuring the velocity of flow in several spots in a stream and multiplying the flow velocity over the cross-sectional area of the stream. However, this method is effort-intensive. An easier way to approximate discharge is by developing a **rating curve** for a stream at a given sampling point. To develop a rating curve, a series of measurements of discharge and stream stage must be made together to develop a mathematical relationship. Then, subsequent discharge measurements can be calculated from a measurement of stream stage, measured by a gage that takes readings remotely. 

```{r}
EnoDischarge.ratingcurve <- readNWISdv(siteNumbers = "02096500",
                     parameterCd = c("00060", "00065"),
                     startDate = "2009-08-01",
                     endDate = "2019-07-31")

names(EnoDischarge.ratingcurve)[4:7] <- c("Discharge", "D.Approval.Code",
                                          "Gage.Height", "GH.Approval.Code")

DvsGHplot <-
  ggplot(EnoDischarge.ratingcurve, aes(x = Discharge, y = Gage.Height)) +
  geom_point(alpha = 0.5) #logarithmic relationship -- positive overall correlation but levels out over time (smaller accel)
#logarithmic relationship -- think of cereal bowl
#NOTE: you could switch the axes 
#this is empirically-determined discharge 
print(DvsGHplot)
```

### Data import
We will again be using the dataRetrieval package, which allows us to access USGS hydrologic data. We will be using the hydrologic data capabilities of this package today, but the package also comes with additional functionality to access water quality data from the Water Quality Portal. 

dataRetrieval Vignette:
https://cran.r-project.org/web/packages/dataRetrieval/vignettes/dataRetrieval.html
#follow link for "vignette" (AKA documentation re: the "package")

The function `whatNWISdata` allows us to access metadata about a site, including what variables have been monitored at that site, start and end dates, and the number of samples for a given variable.
#also can gather data from water quality portal (in addition to NWIS)

We will be analyzing data from the following rivers/streams today: 

* Eno River in North Carolina
* Verde River in Arizona (the largest perennial stream in AZ)
* Bitterroot River in Montana
* Sauk River in Minnesota (Dr. Salk's hometown)
* Nehalem River in Oregon

```{r}
EnoSummary <- whatNWISdata(siteNumbers = "02096500")
VerdeSummary <- whatNWISdata(siteNumbers = "09504000")
BitterrootSummary <- whatNWISdata(siteNumbers = "12344000")
SaukSummary <- whatNWISdata(siteNumbers = "05270500")
NehalemSummary <- whatNWISdata(siteNumbers = "14299800")
```

What data are available from the sites? Pay particular attention to the parameter codes and statistic codes.

>NOTE: there are different amounts of rows / data points per river dataset
For Eno River, there are three rows with same parm_cd that stands for discharge
Next, we go to stat_cd and find three codes ... that stand for max, min, mean
Next, move to period of record 

What is the period of record for mean daily discharge at each of these sites?

> Eno River: 1928 to 2019

> Verde River: 1915 to 2019

> Bitterroot River: 1937 to 2019

> Sauk River: 

> Nehalem River: 

The function `readNWISdv` allows us to access daily hydrologic data, which include discharge, gage height, temperature, precipitation, and pH. The function `readNWISqw` allows us to access water quality data. 

We will access the last 10 years of data for our purposes today. Alternatively,  you could code `startDate = ""` to indicate you want to access all available data for the entire period of record. Then, you could filter out the years you don't want to analyze.

```{r}
# Import data
EnoDischarge <- readNWISdv(siteNumbers = "02096500",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2009-09-01",
                     endDate = "2019-08-31")

VerdeDischarge <- readNWISdv(siteNumbers = "09504000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2009-09-01",
                     endDate = "2019-08-31")

BitterrootDischarge <- readNWISdv(siteNumbers = "12344000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2009-09-01",
                     endDate = "2019-08-31")

SaukDischarge <- readNWISdv(siteNumbers = "05270500",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2009-09-01",
                     endDate = "2019-08-31")

NehalemDischarge <- readNWISdv(siteNumbers = "14299800",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "2009-09-01",
                     endDate = "2019-08-31")
#if you delete StartDate specification, will start from earliest available
# if you delete EndDate specification, will end with latest available
#Data from internet, so may take awhile

#dv stands for "daily values"
#imported as parameter_state codes <-- notation
#A stands for "Approved" and P stands for "Provisional"

```

### Data Wrangling
```{r}

# Renaming columns (one method of multiple)
names(EnoDischarge)[4:5] <- c("Eno.Discharge", "Eno.Approval.Code")
names(VerdeDischarge)[4:5] <- c("Verde.Discharge", "Verde.Approval.Code")
names(BitterrootDischarge)[4:5] <- c("Bitterroot.Discharge", "Bitterroot.Approval.Code")
names(SaukDischarge)[4:5] <- c("Sauk.Discharge", "Sauk.Approval.Code")
names(NehalemDischarge)[4:5] <- c("Nehalem.Discharge", "Nehalem.Approval.Code")

# Remove agency_cd column from each data frame
EnoDischarge <- select(EnoDischarge, -agency_cd)
VerdeDischarge <- select(VerdeDischarge, -agency_cd)
BitterrootDischarge <- select(BitterrootDischarge, -agency_cd)
SaukDischarge <- select(SaukDischarge, -agency_cd)
NehalemDischarge <- select(NehalemDischarge, -agency_cd)

# Join data frames
CombinedDischarge <- full_join(EnoDischarge, VerdeDischarge, by = "Date") %>%
  full_join(., BitterrootDischarge, by = "Date") %>%
  full_join(., SaukDischarge, by = "Date") %>%
  full_join(., NehalemDischarge, by = "Date")

#SELECT allows us to select COLUMNS
#FILTER allows us to select ROWS
```

### Data Visualization
```{r}
# Build a ggplot
EnoPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Eno.Discharge)) +
  geom_line() +
  ggtitle("Eno River") +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())
print(EnoPlot)

VerdePlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Verde.Discharge)) +
  geom_line() +
  ggtitle("Verde River") +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())
print(VerdePlot)

BitterrootPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Bitterroot.Discharge)) +
  geom_line() +
  ggtitle("Bitterroot River") +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())
print(BitterrootPlot)

SaukPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Sauk.Discharge)) +
  geom_line() +
  ggtitle("Sauk River") +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12), 
        axis.text.x = element_blank(), axis.title.x = element_blank())         
print(SaukPlot)

NehalemPlot <- 
  ggplot(CombinedDischarge, aes(x = Date, y = Nehalem.Discharge)) +
  geom_line() +
  ggtitle("Nehalem River") +
  labs(x = "", y = expression("Discharge (ft"^3*"/s)")) + 
  theme(plot.title = element_text(margin = margin(b = -10), size = 12))
print(NehalemPlot)

CombinedPlot <- 
  plot_grid(EnoPlot, VerdePlot, BitterrootPlot, SaukPlot, NehalemPlot, 
            ncol = 1)
print(CombinedPlot)
```

How does the magnitude of discharge compare for these rivers?

>Eno a lot higher than rest

Which of these rivers have a seasonal cycle of discharge? How might you explain the presence and timing of seasonal cycles based on climatic conditions?

>Bitteroot and Nehalem have cyclical data. These are both situated in the U.S. Northwest. Perhaps the discharge is driven by freezing and melting of the riverwater itself, as well as snowmelt from mountains above the rivers. These occurrences change as the seasons progress and change.
Montana sunnier -- maybe absence of rain 
Sauk River most recently shows some erratic patterns -- less and less defined seasonal cycle -- climate change -- warming winters, 
  more erratic precipitation
Verde River in AZ, which is dry = not much discharge 

Hypothetical question: If you wanted to create a ggplot using facets instead of `plot_grid`, how would you go about wrangling your dataset differently? 

>Create a column with river name / site number <-- 'rbind'? <-- same column titles in each data frame is necessary for use of this fxn
Wouldn't have to create 5 separate graphs -- instead would do in one move
input --> "ggplot make discharge by date graph and facet by site number"
wouldn't use the "join" function (from the dplyer package)

## Defining "typical" discharge pattern
Let's take the Bitterroot River, which displays a clear seasonal cycle. Let's define the median discharge on a given day of year and overlay this onto the range experienced. We will take a longer period of record this time.
```{r}  
BitterrootDischarge.Historical <- readNWISdv(siteNumbers = "12344000",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "1969-09-01",
                     endDate = "2019-08-31")
#only using one river, so 50 years of data versus 10 years as before

names(BitterrootDischarge.Historical)[4:5] <- c("Discharge", "Approval.Code")

BitterrootDischarge.Historical <- mutate(BitterrootDischarge.Historical, 
                                         DOY = yday(Date))

BitterrootDischarge.Pattern <- BitterrootDischarge.Historical %>%
  group_by(DOY) %>%
  summarise(Median.Discharge = median(Discharge), 
            Max.Discharge = max(Discharge),
            Min.Discharge = min(Discharge))

BitterrootPatternPlot <- 
  ggplot(BitterrootDischarge.Pattern, aes(x = DOY)) +
  geom_line(aes(y = Median.Discharge)) +
  geom_line(aes(y = Max.Discharge), color = "gray") +
  geom_line(aes(y = Min.Discharge), color = "gray") +  
  labs(x = "Day of Year", y = expression("Discharge (ft"^3*"/s)")) 
print(BitterrootPatternPlot)
#DOY because Julian Date & Ordinal Date are different / there are different kinds of dates
#NOTE: pipes mean "then" in English 
```

Challenge: Edit the code above so that rather than plotting the minimum and maximum discharge, you are plotting the 95 % confidence interval of the discharge on a given date. Hint: look up the formula for confidence interval.
#Choosing statistical representative measures (ex. 95% confidence interval or quantiles -- might more accurately represent spread, whereas min and max are sensitive to outliers)

## Closing Discussion
What are your impressions of seasonal and interannual discharge for our example rivers? How might you go about interpreting discharge data for any given river?
#You would interpret discharge based on time / date 
#Precipitation as joint y-axis
#Air temperature -- evapotranspiration and its effect on flow 
